Agile Formality: A ``Mole'' of Software Engineering Practices

### 2010-08-19 papers/AgileFormality/paper.tex
changes to reinclude ifpdf after bug solved some packages removed
formatting packages removed
small changes and clarifications introduced. latex style cleaned up. to be done: expand gained experience, clarify prerequisites to apply methodology, reduce to 20 pages

%\documentclass{article}
\documentclass[english]{lni}
\usepackage{latin1}

\usepackage{times}
%\usepackage{ifpdf}
\usepackage{a4wide}
\usepackage{pdfsync}

@@

\usepackage{xspace} 
\usepackage{tabularx} 
\usepackage{epsfig}
\usepackage{amsmath} 
\usepackage{amsfonts} 
\usepackage{amssymb}
\usepackage{eucal} 
\usepackage{stmaryrd} 
\usepackage{float}
\usepackage{listings}

@@

Members of the agile programming and formal methods communities do not always see eye-to-eye.  
These two communities often do not talk to or learn from each other.  
Only recently, as highlighted by the September 2009 issue of IEEE Software, the IFIP workshop on balancing agility and formalism in software engineering, and the first edition of the international workshop for formal methods and agile methods, ideas from the two communities begun to be synthesized.  
While the problem-solving approaches and psychological attitudes of members of the two communities differ widely, we exploit this clash of viewpoints, exploiting a new development processes that actually blends, rather than mashes together, best practices from the two worlds.  
This paper summarizes our process and a supporting complex case study, showing that it is not only possible, but tasty, to combine the ``chili pepper'' of formal methods and the ``chocolate'' of agile programming, thus producing a tasty ``Mole'' (as in the highly-spices Mexican sauce) of software engineering practices.

@@

% terminology
In this paper the IEEE 610~/cite{Jay1990} standard terminology for validation and verification is used.  
Paraphrasing the standard slightly, \emph{verification} is the process of evaluating software to determine whether the products of a given development phase satisfy the conditions imposed at the start of that phase, according to the requirements and design specifications (``you build the system in the
right way''); \emph{validation} is the process of evaluating software during or at the end of the development process to determine whether it satisfies specified requirements ensuring it meets the user's needs (``you build the right system'').

Within the formal methods community, verification means something much stronger, and validation something quite different, than in the IEEE 610 standard.
Loosely, within the FM community, \emph{verification} (in the paper \emph{formal verification}) means the use of formal methods to formally, statically (and usually modularly) verify that a system conforms to its specification (under all possible execution scenarios).  
On the other hand, \emph{validation} (in ther paper \emph{formal validation}) means the use of traditional software engineering practices to check, statically or dynamically, that a system conforms to its specification (whether in the form of tests, requirements, etc.) by hand or via execution under some execution scenarios.
In the present paper both perspectives are respected and acknowledged, with the attempt to be absolutely clear in using these terms unambigously.
Additionally, ``...development of a system,'' more precisely means ``...analysis, design, development, verification, formal validation of a system''.

@@

% test driven kinds
The test driven approach relies on writing the tests before the actual implementation. 
It is appliable in a very small scale (Test Driven Development) yielding to activity cycles as short as a few minutes.
It is appliable in a large scale as well (requirements immediately translated into functional tests), yeilding to activity cycles as long as a full delivery iteration (usually no longer than a couple of months).

@@

% previous work: a common approach
As already mentioned, in all these attempts, the handmade test suites are automatically generated using the system formal specification, or completely replaced by theorem provers or static checkers. 
The requisite to enable these approaches is, at least, a specification of the entire system under development to enable simple verification techniques as runtime verification (obtained by assertions and design by contract~\cite{Meyer1997} preconditions, postconditions and invariants) and test suites automatic generation~\cite{Cheon2002,Cheon2004,Cheon2005}.
At most, a complete and sound system specification, usually supported by an appropriate formal language (to specify model and properties) and a constrained programming language, used to actually prove properties about the system under development~\cite{CatanoHuisman02,DetlefsNelsonSaxe2005,KiniryCok04}.

@@

% what we want to achieve
The objectives we want to achieve are the following: to formally specify only parts of the system, in order to cope with constrained resources and unstable requirements; to draft an highly iterative and incremental development process that blends formal and agile practices.

% how we want to achive it
The main specific problem, by now never addressed, is how to connect development artifacts of the two worlds together: we will define activity cycles, similar to what is found in agile methodologies (especially in XP), to solve this problem in an highly iterative development process.
We will show how tests can drive the formal specification, how the formal specifications can drive tests and code development, how handmade and automated tests can coexist and support each other, how the unspecified parts of the system can be incrementally specified.
That is, we will show how to blend, and not compose, formal validation and agile verification practices: a ``Mole'' of verification practices\footnote{A mole (\emph{'m\=ol\=a}) is a highly spiced Mexican sauce made chiefly from chili peppers and chocolate, served with meat.}.

@@

% composition is not enough
Is this enough to deal with formal and informal artifacts?
FDD is used for the formal specified parts of the system, whether TDD is used for the remaining ones: a ``composition'' of formal and agile verification practices.
Yet, it is not enough.
There are several problems that still need to be addressed.
No communication between the two worlds is permitted, it is not defined how to formally specify a module after it has been implemented and verified in a non formal way.
A module only partially specified is a sort of hybrid that must be treated accordingly.
Finally, in the FDD cycle shown in Figure \ref{fig:fdd_diagram}, the implicit assumption is that it is always possible to identify the correct formal specifications of the system, but this is not always straighforward.
Identify the correct formal specification is a difficult task, it has to be supported and verified by the development process.

@@

% verification
Verification is obtained through an handmade test suite, composed of unit and integration tests; the tests specifies the behaviours of the implementation, which is capable of recognizing the proper packet structure in a binary data stream. 
The implementation is tricky: it searches for repeated patterns into a byte stream to be sufficiently confidant on how the stream is decomposed into packets.
Besides, the packet byte structure is not stable: the internal byte structure changed several times during development.
Therefore, packet byte structure is not formally specified.

@@

% test effectiveness
Code coverage metrics are used to compare in a quantitative way the effectiveness of the test suites. 
This is only an indicator, since it is accepted that code coverage alone cannot asses the quality of a test suite\cite{Marick1999,Chockler2006}.
Various coverage metrics exist, the test suites are analyzed with statement code coverage (one of the simplest types)\footnote{The tool used to obtain statement code coverage metrics is \myhref{http://emma.sourceforge.net/}{Emma}.}: statement coverage reports whether each statement is encountered during execution. 
Statement code coverage is easy to implement, but lacks the capability to explore the control flow graph associated to the code, that is, decision branches are not explored. 
The coverage results are reported in the Table \ref{tab:statement_code_coverage}.

@@

% Mole
The verification practices proposed recreate the fast feedback development environment we find in Test Driven Development, used with formal and non formal artifacts.
A total of four practices are presented, two of them are conservative, following closely the development practices of agile and formal methodologies, while the other pair are inoovative, blending together elements from both worlds.

% Mole at work
The practices have been applied succesfully to develop a driver for a custom embedded sensor board, equipped with board simulators and protocol verifiers. 
The protocol and the board specifications were informal and unstable. 
Yet a working software product was needed as soon as possible, with incrementally added functionality, and keeping it linked with the last specifications of the protocol and the board.
``Mole'' verification practices enabled us to keep in sink formal specifications (JML annotations), informal specifications (test suites) and source code, through very fast and short development cycles.
We developed handmade test suites, generated test suites and JML annotations, wich succesfully supported the verification of the driver and the simulators, and the board as well, when it was delivered.

% extending Mole
The practices are very general, and are appliable to different development languages and formal methodologies.
One important constraint is on the formal methodology, that must be able to cope with partial specifications; another constraint is on the tools available, that must be able to verify the system properties at runtime and partially automate the verification process of the formalized properties.

>>>>

%\documentclass[english]{lni}
\documentclass{article}
\usepackage{latin1}

%\usepackage{times}
\usepackage{ifpdf}
%\usepackage{a4wide}
%\usepackage{pdfsync}

@@

\usepackage{xspace} 
%\usepackage{tabularx} 
%\usepackage{epsfig}
%\usepackage{amsmath} 
%\usepackage{amsfonts} 
%\usepackage{amssymb}
%\usepackage{eucal} 
%\usepackage{stmaryrd} 
\usepackage{float}
\usepackage{listings}

@@

Members of the agile programming and formal methods communities do not always see face-to-face.  
These two communities often do not talk to or learn from each other.  
Only recently, as highlighted by the September 2009 issue of IEEE Software, the IFIP workshop on balancing agility and formalism in software engineering, and the first edition of the international workshop for formal methods and agile methods, ideas from the two communities begun to be synthesized.  
While the problem-solving approaches and psychological attitudes of members of the two communities differ widely, we exploit this clash of viewpoints, exploiting a new development processes that actually blends, rather than mashes together, best practices from the two worlds.  
This paper summarizes our process and a supporting complex case study, showing that it is not only possible, but tasty, to combine the ``chili pepper'' of formal methods and the ``chocolate'' of agile programming, thus producing a tasty ``Mole'' (as in the highly-spiced Mexican sauce) of software engineering practices.

@@

% IEEE terminology
In this paper the IEEE 610~/cite{Jay1990} standard terminology for validation and verification is used.  
Paraphrasing the standard slightly, \emph{verification} is the process of evaluating software to determine whether the products of a given development phase satisfy the conditions imposed at the start of that phase, according to the requirements and design specifications (``you build the system in the right way''); \emph{validation} is the process of evaluating software during or at the end of the development process to determine whether it satisfies specified requirements ensuring it meets the user's needs (``you build the right system'').

% FM terminology
Within the formal methods community, verification means something much stronger, and validation something quite different, than in the IEEE 610 standard.
Loosely, within the FM community, \emph{verification} (in the paper \emph{formal verification}) means the use of formal methods to formally, statically verify that a system conforms to its specification.
On the other hand, \emph{validation} (in the paper \emph{formal validation}) means the use of traditional software engineering practices to check, statically or dynamically, that a system conforms to its specification (whether in the form of tests, requirements, etc.) by hand or via execution under some execution scenarios.
In the present paper both perspectives are respected and acknowledged.

% tests terminology
Tests are usually classified by dichotomies: unit vs. functional, black-box vs. white-box, testing vs. design, tester vs. coder. 
Yet testing doesn’t always fit in the dichotomies, especially in the case of agile testing approaches~/cite{Beck2007}.
Black-box vs. white-box dichotomy is particularly problematic, since the box boundaries change the perspective.
In the paper, functional tests are always black-box tests, coverage tests are always white-box tests, generated tests are always black-box tests (the box is the module specified); other strict rules do not hold.

% development terminology
Additionally, ``...development of a system,'' more precisely means ``...analysis, design, development, verification, formal validation of a system''.

@@

% test driven kinds
The test driven approach relies on writing the tests before the actual implementation. 
It is applicable to a very small scale (Test Driven Development) yielding to activity cycles as short as a few minutes.
It is applicable to a large scale as well (requirements immediately translated into functional tests), yielding to activity cycles as long as a full delivery iteration (usually no longer than a couple of months).

@@

% previous work: a common approach
As already mentioned, in all these attempts, the test suites are automatically generated using the system formal specification, or completely replaced by theorem provers or static checkers. 
The requisite to enable these approaches is, at least, a specification of all the relevant aspects of the system under development, to enable simple verification techniques such as runtime verification (obtained by assertions and design by contract~\cite{Meyer1997} preconditions, postconditions and invariants) and test suites automatic generation~\cite{Cheon2002,Cheon2004,Cheon2005}.
At most, a complete and sound system specification, usually supported by an appropriate formal language (to specify model and properties) and a constrained programming language, used to statically prove properties about the system~\cite{CatanoHuisman02,DetlefsNelsonSaxe2005,KiniryCok04}.

@@

% outsider
Liu in~\cite{Liu2010} takes a different approach where SOFL methodology~\cite{Liu2004} is merged into agile processes. 
SOFL methodology is based on a three step specification approach: (1) informal specification, (2) semi-formal specification, (3) formal specification. 
The agile adaptation consists mainly in shorter loops between activities, and into a reduced set of formal specifications.
The formal specifications are used only to help understanding ambiguous statements in the semi-formal specification, and are not maintened while the system evolves; testing and inspections are based on the semi-formal specification.
The formal specification are partial, which is interesting, but used only for documentation purposes and than discarded.

% what we want to achieve
The objectives we want to achieve are the following ones: to formally specify only parts of the system, in order to cope with constrained resources and unstable requirements; to draft an highly iterative and incremental development process that blends formal and agile practices.

% how we want to achive it
The main specific problem, by now never addressed, is how to connect development artifacts of the two worlds together: we will define activity cycles, similar to what is found in agile methodologies (especially in XP), to solve this problem in an highly iterative development process.
We will show how tests can drive the formal specification, how the formal specifications can drive tests and code development, how handmade and automated tests can coexist and support each other, how the unspecified parts of the system can be incrementally specified.
That is, we will show how to blend, and not compose, formal validation and agile verification practices: a ``Mole'' of verification practices\footnote{A mole (\emph{'m\=ol\=a}) is a highly spiced Mexican sauce made chiefly from chili peppers (agile) and chocolate (formal), served with meat (working system).}.

@@

% composition is not enough
Is this enough to deal with formal and informal artifacts?
FDD is used for the formal specified parts of the system, whether TDD is used for the remaining ones: a ``composition'' of formal and agile verification practices.
Yet, it is not enough.
There are several problems that still need to be addressed.
No communication between the two worlds is permitted, it is not defined how to formally specify a module after it has been implemented and verified in a non formal way.
A module only partially specified is a sort of hybrid that must be treated accordingly.
Finally, in the FDD cycle shown in Figure \ref{fig:fdd_diagram}, the implicit assumption is that it is always possible to identify the correct formal specifications of the system, but this is not always straighforward.
Identifying the correct formal specification is a difficult task, it has to be supported and verified by the development process.

@@

% verification
Verification is obtained through an handmade test suite, composed of unit and integration tests; the tests specifies the behaviours of the implementation, which is capable of recognizing the proper packet structure in a binary data stream. 
The implementation is tricky: it searches for repeated patterns into a byte stream to be sufficiently confident on how the stream is decomposed into packets.
Besides, the packet byte structure is not stable: the internal byte structure changed several times during development.
Therefore, packet byte structure is not formally specified.

@@

% test effectiveness
Code coverage metrics (white-box testing) are used to compare in a quantitative way the effectiveness of the test suites. 
This is only an indicator, since it is accepted that code coverage alone cannot asses the quality of a test suite\cite{Marick1999,Chockler2006}.
Various coverage metrics exist, the test suites are analyzed with statement code coverage (one of the simplest types)\footnote{The tool used to obtain statement code coverage metrics is \myhref{http://emma.sourceforge.net/}{Emma}.}: statement coverage reports whether each statement is encountered during execution. 
Statement code coverage is easy to implement, but lacks the capability to explore the control flow graph associated to the code, that is, decision branches are not explored. 
The coverage results are reported in the Table \ref{tab:statement_code_coverage}.

@@

% Mole
The verification practices proposed recreate the fast feedback development environment we find in Test Driven Development, used with formal and non formal artifacts.
A total of four practices are presented, two of them are conservative, following closely the development practices of agile and formal methodologies, while the other pair are innovative, blending together elements from both worlds.

% Mole at work
The practices have been applied succesfully to develop a driver for a custom embedded sensor board, equipped with board simulators and protocol verifiers. 
The protocol and the board specifications were informal and unstable. 
Yet a working software product was needed as soon as possible, with incrementally added functionality, and keeping it linked with the last specifications of the protocol and the board.
``Mole'' verification practices enabled us to keep in sync formal specifications (JML annotations), informal specifications (test suites) and source code, through very fast and short development cycles.
We developed handmade test suites, generated test suites and JML annotations, wich succesfully supported the verification of the driver and the simulators, and the board as well, when it was delivered.

% extending Mole
The practices are very general, and are applicable to different development languages and formal methodologies.
One important constraint is on the formal methodology, that must be able to cope with partial specifications; another constraint is on the tools available, that must be able to verify the system properties at runtime and partially automate the verification process of the formalized properties.

\bibliographystyle{lni}
